---
title: "Multi-State-Model Wave2"
author: "Jonathan Somer"
date: "12/26/2020"
output: html_document
params:
  positive_to_hospitalized_csv: NULL
  hospitalized_onward_csv: NULL
  model_cache_file: NULL
  use_15_9_reduced_model: NULL
---

```{r}
print(params)

POSITIVE_TO_HOSPITALIZED_CSV = params$positive_to_hospitalized_csv # 'data/fit_data/01-positive_to_hospitalized.csv'
HOSITALIZED_ONWARD_CSV = params$hospitalized_onward_csv # 'data/fit_data/02-hospitalized_onward.csv'

MODEL_CACHE_FILE = params$model_cache_file
USE_15_9_REDUCED_MODEL = params$use_15_9_reduced_model

stopifnot((!is.null(POSITIVE_TO_HOSPITALIZED_CSV)) & (!is.null(POSITIVE_TO_HOSPITALIZED_CSV)))
```




```{r}

source('utils/wave2_utils.R')
source('./model/wave2_model.R')
#library(formattable)

VERBOSE = FALSE
```


# Fit is done in three stages: 

1. Fit a `pos-> hosp/home` `CompetingRisksModel()`
2. Fit a `hosp -> ...` `MultiStateModel()`
3. Add the `pos-> hosp/home`model to the multi-state model.

## 1. Fit Competing Risks Model From Positive Diagnosis:

### Columns used: 

- `first_positive_test_date`
- `days_from_positive_to_hosp` (date to censoring in case of no first_date, used as time to `home`)
- `new_type0` (hospitalized at this state)
- `sector` 

** Not used: **
- `first_date` (date of hospitalization)

```{r}
pos_df = read.csv(POSITIVE_TO_HOSPITALIZED_CSV, row.names = 1,stringsAsFactors=FALSE)
pos_df$sector = ifelse(pos_df$sector == "unknown", yes = "arab", no = pos_df$sector)
stopifnot(sum(pos_df$sector == "unknown") == 0)
#formattable(head(pos_df, n=5))
```

```{r}
. = format_positive_to_hospitalization_df(pos_df, use_sector=TRUE)
with_sector_pos_df = .$pos_df
with_sector_covariate_names = .$covariate_names

# make sure using split MnM state
stopifnot(sum(with_sector_pos_df$type0 == MILD_OR_MODERATE) == 0)

stopifnot(max(with_sector_pos_df[with_sector_pos_df$type0==HOME,]$days_from_positive_to_hosp) > 25)

#formattable(head(with_sector_pos_df, n=5))
```


```{r}
t_transition = with_sector_pos_df$days_from_positive_to_hosp
target_state = with_sector_pos_df$type0
covariates_X = with_sector_pos_df[,with_sector_covariate_names]

with_sector_pos_to_hosp_model = CompetingRisksModel()
with_sector_pos_to_hosp_model$fit(t_transition, target_state, covariates_X, verbose = 2)
```

### Sanity Check 1:

```{r}
with_sector_covariates_example = covariates_X[1,]
sum_probs = print_next_state_probabilities(with_sector_pos_to_hosp_model, with_sector_covariates_example, 0)
stopifnot( (sum_probs >= 0.9) && (sum_probs <= 1.1) )
```

### PATCH: only `pos->home` should use sector.


1. Fit `c(age,sex, age*sex)` CompetingRisksModel at state POSITIVE:

```{r}
. = format_positive_to_hospitalization_df(pos_df, use_sector=FALSE)
no_sector_pos_df = .$pos_df
no_sector_covariate_names = .$covariate_names

# make sure using split MnM state
stopifnot(sum(no_sector_pos_df$type0 == MILD_OR_MODERATE) == 0)

#formattable(head(no_sector_pos_df, n=5))
```


```{r}
t_transition = no_sector_pos_df$days_from_positive_to_hosp
target_state = no_sector_pos_df$type0
covariates_X = no_sector_pos_df[,no_sector_covariate_names]

no_sector_pos_to_hosp_model = CompetingRisksModel()
no_sector_pos_to_hosp_model$fit(t_transition, target_state, covariates_X, verbose=2)
```

### Sanity Check 2:

```{r}
no_sector_covariated_example = covariates_X[1,]
sum_probs = print_next_state_probabilities(no_sector_pos_to_hosp_model, no_sector_covariated_example, 0)
stopifnot( (sum_probs >= 0.9) && (sum_probs <= 1.1) )
```

Extend no-sector model covariates to allow for receiving samples with all covariates:

NOTE! - In order to use the correct baseline hazard function, we need to stretch the covariates of the age,sex only model and use that model, AND DO NOT overwrite the covariates of the with_sector_ model (whose baseline hazard function corresponds to the baseline of "all-zeros" patients, in this case: an aged 0, female, **arab** patient)

```{r}

for (f in no_sector_pos_to_hosp_model$failure_types) {
  coefs = no_sector_pos_to_hosp_model$event_specific_models[[f]]$coefficients
  
  new_coefs = c(
    # 1. age
    # 2. gendermale
    coefs[c(1,2)], 
    
    # 3. sectorharedim
    # 4. sectormixed
    # 5. sectorrest
    rep(0, 3),
    # 6. age:gendermale
    coefs[c(3)],
    # 7. age:sectorharedim
    # 8. age:sectormixed
    # 9. age:sectorrest
    # 10. gendermale:sectorharedim
    # 11. gendermale:sectormixed
    # 12. gendermale:sectorrest
    rep(0, 6)
  )

  no_sector_pos_to_hosp_model$event_specific_models[[f]]$coefficients = new_coefs
}
```


Overwrite `pos->home` model in the `no_sector_pos_to_hosp_model` CompetingRisksModel() with one that utilizes sector.


```{r}
no_sector_pos_to_hosp_model$event_specific_models[[HOME]] = with_sector_pos_to_hosp_model$event_specific_models[[HOME]]
```


Resulting coefficients of all models:

```{r}

coefficients_table = function(cr_model, covariate_names) {
  
  coefficients_df = data.frame()
  for (next_state in 1:length((cr_model$event_specific_models))) {
    
    model = cr_model$event_specific_models[[next_state]]
    
    if (is.null(model)) next
    
    coefficients_df = rbind(coefficients_df, c(next_state, model$coefficients))
  }
  
  colnames(coefficients_df) = c("next_state", covariate_names)
  return(coefficients_df)
}

#formattable(coefficients_table(no_sector_pos_to_hosp_model, with_sector_covariate_names))
```

Rename the "no_sector" CompetingRisksModel which now allows sector covariates, and includes the pos->home model which has sector covariates

```{r}
pos_to_hosp_model = no_sector_pos_to_hosp_model
```


### Sanity Check 3:
```{r}
sum_probs = print_next_state_probabilities(pos_to_hosp_model, with_sector_covariates_example, 0)
stopifnot( (sum_probs >= 0.9) && (sum_probs <= 1.1) )
```



## 2. Fit "hosp -> ... " Multi-State-Model: 


```{r}
hosp_df = read.csv(HOSITALIZED_ONWARD_CSV,stringsAsFactors=FALSE)
#formattable(head(hosp_df, n=5))
```


```{r}
. = construct_msm_dataset_from_hospitalized_onward_df(hosp_df)
dataset = .$dataset
covariate_names = .$covariate_names

dataset[[1]]
```


```{r}

# sample_covariates on entry to origin_state
# target_state might be MILD / MODERATE if leaving POSITIVE state.
update_covariates_function = function(sample_covariates, origin_state, target_state, time_at_origin_state, absolute_time_of_entry_to_target_state=NULL) {
  
  # this shouldn't happen:
  if ( (origin_state == POSITIVE) & (target_state == HOME) ) NULL$zappa
  
  if (origin_state == POSITIVE) {
    # # can assume patient has "positive-style" covariates:
    
    return(
      construct_hospitalized_patient_covariates(
        age = sample_covariates[AGE_COVARIATE_IDX],
        is_male = sample_covariates[IS_MALE_COVARIATE_IDX],
        state_at_hospitalization = target_state,
        cumulative_time_since_hospitalization = 0,
        was_severe = 0,
        was_critical = 0
      )
    )
    
  } else { 
    # can assume patient has "hospitalized-style" covariates:
    
    return(
      construct_hospitalized_patient_covariates(
        age = sample_covariates[AGE_COVARIATE_IDX],
        is_male = sample_covariates[IS_MALE_COVARIATE_IDX],
        state_at_hospitalization = get_state_at_hospitalization(sample_covariates),
        
        # in case prediction starts from t > 0. The cumulative time can be 0, as it was in entry to state.  
        # But the cumulative time in the next state needs to include both the time spent in the state thus far, and the remaining time
        cumulative_time_since_hospitalization = max(sample_covariates[CUMULATIVE_TIME_COVARIATE_IDX] + time_at_origin_state, 
                                                    absolute_time_of_entry_to_target_state), 
        
        was_severe = sample_covariates[WAS_SEVERE_COVARIATE_IDX] | (origin_state == SEVERE),
        was_critical = sample_covariates[WAS_CRITICAL_COVARIATE_IDX] | (origin_state == CRITICAL)
      )
      
    )    
  }
}


model = Wave2MultiStateModel()
model$fit(dataset,
          terminal_states = TERMINAL_STATES,
          update_covariates_function,
          covariate_names=covariate_names)
```

```{r}
comp_ds = model$prepare_dataset_for_competing_risks_fit(dataset, TERMINAL_STATES, update_covariates_function)
crit_to_disch_df = comp_ds[comp_ds$origin_state == 5 & comp_ds$target_state == 7,5:20]
sum(crit_to_disch_df$state_at_hospitalizationmild)
sum(crit_to_disch_df$state_at_hospitalizationmoderate)
sum(crit_to_disch_df$state_at_hospitalizationsevere)
# then 15 hospitalized as critical
```

# Overwrite 2 transitions with models that have reduced covariates:

```{r}

# fit entire multistate model with reduced covariates: 
. = construct_msm_dataset_from_hospitalized_onward_df(hosp_df, use_reduced_covariates = TRUE)
reduced_dataset = .$dataset
reduced_covariate_names = .$covariate_names

# no 
# 1. was_critical
# 2. age:was_critical 
# 3. critical,severe state at hospitalization replaced with merged 
# 4. critical,severe state at hospitalization interaction with age replaced with merged 
length(dataset[[1]]$covariates) - length(reduced_dataset[[1]]$covariates)
```



```{r}
update_REDUCED_covariates_function = function(sample_covariates, origin_state, target_state, time_at_origin_state, absolute_time_of_entry_to_target_state=NULL) {
    return(
      construct_hospitalized_patient_REDUCED_covariates(
        age = sample_covariates[REDUCED_AGE_COVARIATE_IDX],
        is_male = sample_covariates[REDUCED_IS_MALE_COVARIATE_IDX],
        state_at_hospitalization = get_REDUCED_state_at_hospitalization(sample_covariates),
        
        # in case prediction starts from t > 0. The cumulative time can be 0, as it was in entry to state.  
        # But the cumulative time in the next state needs to include both the time spent in the state thus far, and the remaining time
        cumulative_time_since_hospitalization = max(sample_covariates[REDUCED_CUMULATIVE_TIME_COVARIATE_IDX] + time_at_origin_state, 
                                                    absolute_time_of_entry_to_target_state), 
        
        was_severe = sample_covariates[REDUCED_WAS_SEVERE_COVARIATE_IDX] | (origin_state == SEVERE)
      )
    )    
  }
```


```{r}
reduced_model = Wave2MultiStateModel()
reduced_model$fit(reduced_dataset,
                  terminal_states = TERMINAL_STATES,
                  update_REDUCED_covariates_function,
                  covariate_names=reduced_covariate_names)
```


```{r}

stretch = function(reduced_coefs) {
  
    new_coefs = c(
    # 1.age                                  
    # 2.sexmale                              
    # 3.state_at_hospitalizationmild         
    # 4.state_at_hospitalizationmoderate    
    reduced_coefs[c(REDUCED_AGE_COVARIATE_IDX,
                    REDUCED_IS_MALE_COVARIATE_IDX,
                    REDUCED_MILD_COVARIATE_IDX,
                    REDUCED_MODERATE_COVARIATE_IDX)],
    
    # 5.state_at_hospitalizationsevere
    c(0),
   
    # 6.cumulative_time                      
    # 7.was_severe                           
    reduced_coefs[c(REDUCED_CUMULATIVE_TIME_COVARIATE_IDX,
                    REDUCED_WAS_SEVERE_COVARIATE_IDX)],
    
    # 8.was_critical                        
    c(0),
    
    # 9.age:sexmale  
    # 10.age:state_at_hospitalizationmild     
    # 11.age:state_at_hospitalizationmoderate 
    reduced_coefs[c(REDUCED_AGE_IS_MALE_INTERACTION_IDX,
                    REDUCED_AGE_MILD_INTERACTION_IDX,
                    REDUCED_AGE_MODERATE_INTERACTION_IDX)],

    # 12.age:state_at_hospitalizationsevere  
    c(0),
    
    # 13.age:cumulative_time                  
    # 14.age:was_severe                       
    reduced_coefs[c(REDUCED_AGE_CUMULATIVE_TIME_INTERACTION_IDX,
                    REDUCED_AGE_WAS_SEVERE_INTERACTION_IDX)],
    
    # 15.age:was_critical                    
    c(0)
    )
}

# MnM -> Deceased
reduced_MnM_to_Deceased = reduced_model$state_specific_models[[MILD_OR_MODERATE]]$event_specific_models[[DECEASED]]
reduced_MnM_to_Deceased$coefficients = stretch(reduced_MnM_to_Deceased$coefficients)

model$state_specific_models[[MILD_OR_MODERATE]]$event_specific_models[[DECEASED]] = reduced_MnM_to_Deceased

# Discharged -> Deceased
reduced_Discharged_to_Deceased = reduced_model$state_specific_models[[DISCHARGED]]$event_specific_models[[DECEASED]]
reduced_Discharged_to_Deceased$coefficients = stretch(reduced_Discharged_to_Deceased$coefficients)

model$state_specific_models[[DISCHARGED]]$event_specific_models[[DECEASED]] = reduced_Discharged_to_Deceased

if (USE_15_9_REDUCED_MODEL) {
  # Discharged -> Severe
  reduced_Discharged_to_Severe = reduced_model$state_specific_models[[DISCHARGED]]$event_specific_models[[SEVERE]]
  reduced_Discharged_to_Severe$coefficients = stretch(reduced_Discharged_to_Severe$coefficients)

  model$state_specific_models[[DISCHARGED]]$event_specific_models[[SEVERE]] = reduced_Discharged_to_Severe

  ## Severe -> Deceased
  # reduced_Severe_to_Deceased = reduced_model$state_specific_models[[SEVERE]]$event_specific_models[[DECEASED]]
  # reduced_Severe_to_Deceased$coefficients = stretch(reduced_Severe_to_Deceased$coefficients)

  # model$state_specific_models[[SEVERE]]$event_specific_models[[DECEASED]] = reduced_Severe_to_Deceased
}

```



```{r}
#formattable(coefficients_table(model$state_specific_models[[MILD_OR_MODERATE]], covariate_names))
#formattable(coefficients_table(model$state_specific_models[[DISCHARGED]], covariate_names))
#formattable(coefficients_table(model$state_specific_models[[SEVERE]], covariate_names))
```

## Further reduction for 5 -> 7 / Critical -> Discharged Transition

```{r}
# fit entire multistate model with FURTHER reduced covariates:
. = construct_msm_dataset_from_hospitalized_onward_df(hosp_df, use_reduced_covariates = TRUE, critical_state_at_hospitalization_only = TRUE)
further_reduced_dataset = .$dataset
further_reduced_covariate_names = .$covariate_names
# no
# 1. was_critical (-1)
# 2. age:was_critical (-1)
# 3. mild, moderate, severe state at hospitalization replaced with merged "not_critical" (-2)
# 4. mild, moderate, severe state at hospitalization interaction with age replaced with merged (-2)
# total: -6
length(dataset[[1]]$covariates) - length(further_reduced_dataset[[1]]$covariates)
```

```{r}
update_FURTHER_REDUCED_covariates_function = function(sample_covariates, origin_state, target_state, time_at_origin_state, absolute_time_of_entry_to_target_state=NULL) {
    return(
      construct_hospitalized_patient_FURTHER_REDUCED_covariates(
        age = sample_covariates[FURTHER_REDUCED_AGE_COVARIATE_IDX],
        is_male = sample_covariates[FURTHER_REDUCED_IS_MALE_COVARIATE_IDX],
        not_hospitalized_critical = sample_covariates[[FURTHER_REDUCED_NOT_HOSPITALIZED_CRITICAL_COVARIATE_IDX]],

        # in case prediction starts from t > 0. The cumulative time can be 0, as it was in entry to state.
        # But the cumulative time in the next state needs to include both the time spent in the state thus far, and the remaining time
        cumulative_time_since_hospitalization = max(sample_covariates[FURTHER_REDUCED_CUMULATIVE_TIME_COVARIATE_IDX] + time_at_origin_state,
                                                    absolute_time_of_entry_to_target_state),

        was_severe = sample_covariates[FURTHER_REDUCED_WAS_SEVERE_COVARIATE_IDX] | (origin_state == SEVERE)
      )
    )
  }
```

```{r}
# further_reduced_model = Wave2MultiStateModel()
# further_reduced_model$fit(further_reduced_dataset,
#                          terminal_states = TERMINAL_STATES,
#                          update_FURTHER_REDUCED_covariates_function,
#                          covariate_names=further_reduced_covariate_names)
```


```{r}
# reduced_coefs here are "further_reduced" coefficients
further_stretch = function(reduced_coefs) {

    new_coefs = c(
    # 1.age
    # 2.sexmale
    # 3.state_at_hospitalizationmild
    # 4.state_at_hospitalizationmoderate
    # 5.state_at_hospitalizationsevere
    reduced_coefs[c(FURTHER_REDUCED_AGE_COVARIATE_IDX,
                    FURTHER_REDUCED_IS_MALE_COVARIATE_IDX,
                    FURTHER_REDUCED_NOT_HOSPITALIZED_CRITICAL_COVARIATE_IDX,
                    FURTHER_REDUCED_NOT_HOSPITALIZED_CRITICAL_COVARIATE_IDX,
                    FURTHER_REDUCED_NOT_HOSPITALIZED_CRITICAL_COVARIATE_IDX)],

    # 6.cumulative_time
    # 7.was_severe
    reduced_coefs[c(FURTHER_REDUCED_CUMULATIVE_TIME_COVARIATE_IDX,
                    FURTHER_REDUCED_WAS_SEVERE_COVARIATE_IDX)],

    # 8.was_critical
    c(0),

    # 9.age:sexmale
    # 10.age:state_at_hospitalizationmild
    # 11.age:state_at_hospitalizationmoderate
    # 12.age:state_at_hospitalizationsevere
    reduced_coefs[c(FURTHER_REDUCED_AGE_IS_MALE_INTERACTION_IDX,
                    FURTHER_REDUCED_AGE_NOT_HOSPITALIZED_CRITICAL_INTERACTION_IDX,
                    FURTHER_REDUCED_AGE_NOT_HOSPITALIZED_CRITICAL_INTERACTION_IDX,
                    FURTHER_REDUCED_AGE_NOT_HOSPITALIZED_CRITICAL_INTERACTION_IDX)],

    # 13.age:cumulative_time
    # 14.age:was_severe
    reduced_coefs[c(FURTHER_REDUCED_AGE_CUMULATIVE_TIME_INTERACTION_IDX,
                    FURTHER_REDUCED_AGE_WAS_SEVERE_INTERACTION_IDX)],

    # 15.age:was_critical
    c(0)
    )
}
# if (USE_15_9_REDUCED_MODEL) {
#  # Critical -> Discharged
#  further_reduced_Critical_to_Discharged = further_reduced_model$state_specific_models[[CRITICAL]]$event_specific_models[[DISCHARGED]]
#  further_reduced_Critical_to_Discharged$coefficients = further_stretch(further_reduced_Critical_to_Discharged$coefficients)

#  model$state_specific_models[[CRITICAL]]$event_specific_models[[DISCHARGED]] = further_reduced_Critical_to_Discharged
# }
```

```{r}
#formattable(coefficients_table(model$state_specific_models[[CRITICAL]], covariate_names))
```


## 3. Add the `pos-> hosp/home` model to the multi-state model.

```{r}
model$state_specific_models[[POSITIVE]] = pos_to_hosp_model
```


## 3. Cache Final Model:


```{r}
# cache_file_name = paste("./cache/wave2_fitted_model_TIME=", as.character(Sys.time()), ".Rds", sep="")
saveRDS(model, MODEL_CACHE_FILE)

rm(list = rm(list=setdiff(ls(), "MODEL_CACHE_FILE")))

m = readRDS(MODEL_CACHE_FILE)
m$state_specific_models[[1]]$event_specific_models[[2]]$coefficient
```








